Timing and Likelihood of the Origin of Life Derived from Post-Impact Highly Reducing Atmospheres

### 2023-05-16 ms.tex
Davids first round edits

% Title
\title{The Timing and Likelihood of the Origin of Life from Earth's Impact History}

@@

\begin{abstract}
Big impacts on the early Earth would have created highly reducing atmospheres that generated molecules needed for the origin of life, such as nitriles. However, such impactors can be followed by collisions that were still sufficiently big to vaporize the ocean and destroy any incipient life. In this scenario, a post-impact reducing atmosphere needs to be followed by a lack of subsequent sterilizing impacts for life to persist. Using what we know about the impact history of early Earth and what is needed to generate post-impact highly reducing atmospheres, we show that  the median timing of biopoesis in this impact-driven scenario was $\sim 4.35$ Ga. However, uncertainties are large because impact bombardment is inherently stochastic, so biopoesis could have occurred in $\sim 0.5$ billion year window from 4.45 to 3.9 Ga within 95\% uncertainties of our model. Claims of a far narrower window for the origin of life are unsupported. In our optimistic scenario for life starting from post-impact reducing atmospheres, we find that biopoesis is possible in 92\% of stochastic impact realizations. This potentially bodes well for life on rocky planets elsewhere because they will have experienced an early episode of enhanced impact bombardment given how planets form.
\end{abstract}

@@

\citet{Benner_2020} argued that the most likely time for the origin of life in an RNA-world scenario would have been $4.36 \pm 0.1$ Ga in the wake of a $\sim 2 \times 10^{22}$ kg ($\sim 2100$ km) asteroid impact which they called Moneta. They suggested that iron delivered by Moneta's core would have reacted with impact-vaporized ocean water to generate a reducing Hadean atmosphere conducive to the photochemical generation of essential prebiotic nitriles, like HCN and HCCCN. Nitriles are required in prebiotic schemes that synthesize ribonucleotide precursors to RNA \citep{Patel_2015,Powner_2009,Becker_2019}. A transient period of rampant nitrile production after a Moneta impact could have been a window of opportunity for the origin of RNA and, ultimately, life.

A single Moneta impact could also explain the highly-siderophile elements (i.e. iron-loving, abbreviated HSEs) in the Moon's and the Earth's mantles. During the moon-forming impact, most all HSEs should have been sequestered in each planetary core, so extant mantle HSEs are commonly interpreted as evidence for late-accretion impactors. Earth's mantle has substantially more HSEs than the Moon by an amount that cannot be accounted for by Earth's greater gravitation cross section \citep{Day_2015}. A single massive Moneta impact could explain the Earth-Moon HSE discrepancy because, by the statistics of small numbers, Moneta could have missed the moon and hit the Earth \citep{Bottke_2010}.

However, the lunar HSE depletion has explanations that do not require a Moneta impact. Lunar HSEs could have been lost to space during impact-delivery because of the Moon's small gravity \citep{Kraus_2015}. Alternatively, HSEs delivered to the Moon during its $\sim 150$ million-year magma ocean could have been sequestered in the core due to iron sulfide exsolution \citep{Morbidelli_2018,Rubie_2016}. Finally, there is some chance that the Earth's HSEs do not record late impacts because the Moon-forming impact delivered HSEs \citep{Sleep_2016}, or because HSEs are gradual core contributions over time from mantle plumes \citep{Halliday_2023,Mundl_2020}. If indeed Earth's HSEs reflect asteroid bombardment after the Moon formed, then the HSEs can be explained by multiple $\sim 500$ to 2000 km impacts rather than a single big ($\sim 2100$ km) collision.

Recently, \citep{Wogan_2023} use photochemical models of post-impact atmospheres to show that impacts significantly smaller than Moneta can efficiently produce prebiotic molecules. In their simulations, impactor iron equilibrates with vaporized ocean water to generate atmospheric H$_2$, and CH$_4$ and NH$_3$ form thermochemically as the reducing steam atmosphere cools. Once steam condense to an ocean, subsequent photochemistry of the Titan-like atmosphere produces prebiotic nitriles. Modeling shows that the production of both HCN and HCCCN only occurs when $\mathrm{CH_4} / \mathrm{CO_2} \gtrsim 0.1$ while the atmosphere is hazy. In an optimistic scenario, which includes nickel-catalyzed methane production, modeling suggests that $\mathrm{CH_4} / \mathrm{CO_2} > 0.1$ (i.e. significant nitriles) occurs for impacts $> 4 \times 10^{20}$ kg. In a pessimistic modeling case, which assumes only a fraction of impactor iron reacts with the atmosphere \citep{Citron_2022}, a $> 5 \times 10^{21}$ kg impactor is required to produce substantial HCN and HCCCN. Regardless of uncertainty, \citet{Wogan_2023} show that an impactor much smaller than Moneta could deliver the nitriles needed for an RNA origin of life, and that the \citet{Benner_2020} estimated timing for biopoesis could be inaccurate because it is based on Moneta.

Here, we use the \citet{Wogan_2023} results, along with Monte-Carlo simulations of Earth's impact history to make an alternative estimate for when life most likely emerged during an RNA-first process. Our calculations account for the possibility of planet sterilization by impacts that vaporize the ocean \citep{Sleep_1989}. We assume that a life-starting impact is one that produces significant prebiotic nitriles and is not subsequently followed by an ocean-vaporizing impact that destroys the biosphere. By considering the fraction of stochastic impact realizations that do not have an life-starting impact, we also estimate the probability of life beginning if Earth's clock was rerun.

@@

We assume the size-frequency distribution of impactors is identical to the main-belt asteroids \citep[Extended Data Figure 1,][]{Marchi_2014}. Data for the frequency of main-belt asteroids only extends down to about 1 km diameter ($1.3 \times 10^{12} $ kg). To extend the distribution to smaller objects, we use the observed 1400 ratio between the frequency of $> 1$ km and $> 20$ km craters on the Moon. \citet{Morbidelli_2018} uses crater scaling relations to show that 1 km and 20 km craters on the Moon corresponds to 50 m and 1 km asteroids, respectively. The largest main belt asteroid is $\sim 1000$ km, so we log-linearly extrapolate to larger impactors (Figure \ref{fig:sfd_and_flux}). Finally, we normalize the size-frequency distribution to the impact mass required to make a 1 km crater on the moon (50 m object, $m_0 = 1.64 \times 10^{8}$ kg). Figure \ref{fig:sfd_and_flux}a shows the resulting size-frequency distribution.

\begin{figure}
  \centering
  \includegraphics[width=1.0\textwidth]{figures/SFD_and_flux.pdf}
  \caption{Our assumed (a) size-frequency distribution of impactors and (b) lunar cratering record. Most of the size-frequency distribution is identical to the main-belt asteroids \citep[Extended Data Figure 1,][]{Marchi_2014}. We extrapolate to objects $\lesssim 10^{11}$ kg using the observed frequency ratio of $> 1$ km and $> 20$ km lunar craters, and log-linearly extrapolate to asteroids $\gtrsim 10^{21}$ kg. The lunar cratering record is the red line in Figure 5 of \citet{Morbidelli_2018}.}
  \label{fig:sfd_and_flux}
\end{figure}

The flux of impactors, $F_0(t)$, can be estimated from the lunar cratering record. We adopt the accretionary tail scenario discussed in \citet{Morbidelli_2018}. In other words, we assume that the ``late heavy bombardment'' did not happen \citep{Cartwright_2022,Hartmann_2019,Zellner_2017}, and that the impact flux on Earth was constantly decreasing throughout the Hadean eon. Figure \ref{fig:sfd_and_flux}b shows $n_m$, our assumed lunar impact history taken from \citet{Morbidelli_2018} (red line in their Figure 5).

Extrapolating the lunar impact history to Earth requires correcting for Earth's greater gravitational attraction. Assuming the approach velocity of impactors far from the Earth and Moon was on average 18 km/s \citep{Morbidelli_2018}, then Earth should receive $s_0 = 1.36$ times more impacts per surface area than the Moon. Therefore, the number of impacts on Earth is

\begin{equation}
  N_0 = A_\oplus s_0 n_m = A_E s_0 10^{a e^{b t} + c}
  \label{eq:num_earth_impacts}
\end{equation}
Equation \eqref{eq:num_earth_impacts} also accounts for Earth's surface area ($A_\oplus$), making $N_0$ the number of impacts on Earth since age $t$ that cause a crater $> 1$ km on the moon. As discussed previously, \citet{Morbidelli_2018} use crater scaling relations to show that a 1 km lunar crater corresponds to a 50 m object with mass $m_0 = 1.64 \times 10^8$ kg.

The derivative of $N_0$ gives the flux, $F_0$:

@@

To simulate impact histories, we consider a fine grid of impactor masses between $10^{12}$ and $10^{23}$ kg, and a grid of times between 4.5 and 3.5 Ga. Indexes $j$ and $i$ indicate the mass and time grid cells respectively, while, for example, $j-\frac{1}{2}$ and $j+\frac{1}{2}$ indicate the edges of the grid cell. We can compute the expected number of impacts within a mass and time grid cell with the following

\begin{equation}
  \overline{N}_{ij} = N(m_{j-\frac{1}{2}},t_{i-\frac{1}{2}},t_{i+\frac{1}{2}}) - N(m_{j+\frac{1}{2}},t_{i-\frac{1}{2}},t_{i+\frac{1}{2}})
\end{equation}
Next, we sample a Poisson distribution for each $\overline{N}_{ij}$, giving a stochastic number of impacts in each mass and time grid cell which constitutes an impact history. Performing this sampling thousands of times captures many of the possible impact histories. We only keep impact histories that accrete a total mass of $2 \times 10^{22}$ and $6 \times 10^{22}$ kg because this is approximately the mass implied by the HSEs in Earth's mantle. This range of masses is based on \citet{Day_2015} who used mantle HSEs to suggest that the Earth was impacted by $\sim 3 \times 10^{22}$ to $4.8 \times 10^{22}$ kg of material, but we choose a wider range of masses because HSEs could have been lost to Earth's core or to space during impacts \citep{Marchi_2018}.

Our final step is to assign impact velocities to each collision in the many sampled impact histories. Appendix Figure \ref{fig:velocity_distribution} shows Earth's impact velocity distribution. We created this distribution using the JPL database of close approaches to Earth by small bodies \citep{Park_2023}, considering all close approach asteroids within 0.05 AU to Earth. The database gives the approach velocity of each asteroid far from Earth, so we compute the impact velocity by accounting for Earth's gravitational energy.

The final collection of sampled impact histories can then be used to compute impact statistics which are relevant to the timing and likelihood of the origin of life.

@@

Moneta also occurs in the Figure \ref{fig:example_impact_histories}b impact realization. In this scenario, any life created in the wake of Moneta should not be subsequently exterminated because Moneta is the last impact to vaporize the ocean.

Figure \ref{fig:example_impact_histories}c shows a impact history where Moneta does not occur. Instead, Earth's mantle HSEs are delivered by multiple $10^{21}$ to $4 \times 10^{21}$ kg collisions. Whether this bombardment history would cause biopoesis depends on the required minimum impact mass to produce significant origin of life nitriles (e.g. HCN and HCCCN). \citet{Wogan_2023} uses photochemical models of post-impact atmospheres to show that, optimistically, an impact $> 4 \times 10^{20}$ kg should give rise to an atmosphere that makes significant HCN and HCCCN. Under this optimistic requirement, the last ocean-vaporizing impact in Figure \ref{fig:example_impact_histories}c at 4.38 Ga with mass $3 \times 10^{21}$ kg could successfully start life. However, with pessimistic modeling assumptions, \citet{Wogan_2023} finds that a $> 5 \times 10^{21}$ kg impact is instead required for substantial nitrile delivery to the surface. In this alternative scenario, an origin of life never occurs in Figure \ref{fig:example_impact_histories}c because there is no impact larger than $5 \times 10^{21}$ kg.

@@

As illustrated in Figure \ref{fig:example_impact_histories}, a life-starting impact does not occur in every simulated impact history. In our impact-driven model for the origin of life, biopoesis requires (1) an impact of sufficient mass to make prebiotic nitriles and (2) the lack of a later ocean-vaporizing collision that destroys the biosphere without rekindling it. Figure \ref{fig:probabilities_of_impacts} shows the probability of conditions (1) and (2) occurring as a function of the minimum impact mass that produces prebiotic molecules. For example, consider the \citet{Wogan_2023} optimistic minimum mass of $4 \times 10^{20}$ kg indicated in Figure \ref{fig:probabilities_of_impacts}. All simulated impact histories have at least one impact bigger than $4 \times 10^{20}$ kg (blue line in Figure \ref{fig:probabilities_of_impacts}). However, the probability of the origin of life in this scenario is 92\%, because 8\% of the time the last $4 \times 10^{20}$ kg collision is followed by a smaller ocean-vaporizing impact that perhaps sterilizes the planet (orange dashed line in Figure \ref{fig:probabilities_of_impacts}). An origin of life is less probable if we consider the \citet{Wogan_2023} pessimistic minimum mass ($> 5 \times 10^{21}$ kg) to produce significant origin of life nitriles. A $> 5 \times 10^{21}$ kg impact without later ocean vaporization occurs 41\% of the time. A life-starting impact was therefore relatively likely on the early Earth, and was strongly favored (92\% probability) under optimistic assumptions.

\begin{figure}
  \centering
  \includegraphics[width=0.85\textwidth]{figures/timing_and_mass.pdf}
  \caption{The timing and mass of a life-starting impact on the early Earth. (a) considers an optimistic minimum impact mass needed to generate the molecules for an RNA origin of life and (b) examines a pessimistic minimum impact mass \citep{Wogan_2023}. In either (a) or (b), panels (i) and (ii) show probability distributions for the timing and mass, respectively, of the last impact the produces prebiotic molecules which does not have a later ocean-vaporizing collision. The vertical dashed lines on (a) i and (b) i plot are median values, and the vertical dotted lines are the 95\% confidence intervals. In an impact-driven origin of life, biopoesis most likely occurred $\sim 4.35$ Ga with a 95\% uncertainty between approximately 3.9 and 4.45 Ga.}
  \label{fig:time_and_mass}
\end{figure}

To estimate the most likely timing for the origin of life, we consider both the \citet{Wogan_2023} optimistic and pessimistic criteria. Figure \ref{fig:time_and_mass}a is the optimistic case, showing the timing (Figure \ref{fig:time_and_mass}a (i)) and mass (Figure \ref{fig:time_and_mass}a (ii)) of the last $> 4 \times 10^{20}$ kg impact for the 92\% of impact histories that do not experience later ocean vaporization. The median timing of a life-starting impact is 4.34 Ga with 95\% uncertainty between 3.89 and 4.46 Ga. The large uncertainty mostly stems from the stochastic, Poisson nature of asteroid bombardment. The probability distribution for the mass of the life-starting impact peaks at $4 \times 10^{20}$ kg and decreases with increasing mass because of the size-frequency distribution of asteroids (Figure \ref{fig:sfd_and_flux}a). There is only a 10\% chance that the impact is Moneta-sized (i.e., between $2 \times 10^{22}$ and $6 \times 10^{22}$ kg).

@@

In this model for biopoesis, life would start approximately 10s of millions of years (i.e., 0.01s of billions of years) after a life-starting impact because this is the maximum duration of the H$_2$- and CH$_4$- rich atmosphere that makes nitriles like HCN \citep{Wogan_2023}. We ignore the $< 10$s million year span between an impact and life's origin because it is small compared to our $\sim 500$ million-year uncertainty for the timing of a life-starting impact.

@@

Our estimated timing for biopoesis is compatible with the earliest well-accepted geologic evidence of life in the 3.5 Ga Pilbara block of western Australia \citep{Walter_1980}. Older geologic evidence of life exists, but is inconclusive. For example, \citet{Bell_2015} discovered graphite inclusions in a 4.1 Ga zircon grain with $^{13}$C depletions compatible with biological activity, but this is a tentative biosignature because there are abiotic mechanisms for making isotopically light carbon \citep{Javaux_2019}. If the \citet{Bell_2015} finding is indeed a sign of life, then it would be mostly compatible with our estimating timing in Figure \ref{fig:time_and_mass}, because we find that there is only a $\sim 10\%$ chance of life starting after 4.1 Ga.

It is temping to extrapolate our results beyond the early Earth to exoplanets, however, we must do so with caution because planet's orbiting different stars may have bombardment histories unlike the Hadean. Planets in the habitable zone of an M-type star during the stellar main-sequence phase were interior to the habitable zone for several hundred million years during the super luminous pre-main-sequence phase \citep{Luger_2015}. \citet{Lichtenberg_2022} uses N-body simulations to show that, for planets hosted by M-dwarfs, large asteroid impacts are not useful for prebiotic chemistry because big impacts most likely occur within 100 million years of planet formation during the stellar pre-main-sequence when the planet is outside of the habitable zone. The pre-main-sequence phase is not an issue for an impact-induced origin of life on planets orbiting sun-like stars because the phase only lasts 10s Myrs, while large asteroid impacts occur for 100s Myrs \citep{Lichtenberg_2022}. Therefore, our result that a life-starting impact was likely on the early Earth (92\% chance in an optimistic case) might be best extrapolated to habitable zone exoplanets orbiting sun-like stars.

>>>>

% Title
\title{Timing and Likelihood of the Origin of Life Derived from Post-Impact Highly Reducing Atmospheres}

@@

\begin{abstract}
Big impacts on the early Earth would have created highly reducing atmospheres that generated molecules needed for the origin of life, such as nitriles. However, such impactors can be followed by collisions that were still sufficiently big to vaporize the ocean and destroy any pre-existing life. Thus, a post-impact reducing atmosphere that gives rise to life needs to be followed by a lack of subsequent sterilizing impacts for life to persist. Using statistics and limits for the impact history on early Earth and the impact mass needed to generate post-impact highly reducing atmospheres, we show that the median timing of impact-driven biopoesis was $\sim 4.35$ Ga. However, uncertainties are large because impact bombardment is stochastic, so biopoesis could have occurred in $\sim 0.5$ billion year window from 4.45 to 3.9 Ga within 95\% uncertainty. Previous claims of a far narrower window for the origin of life are unsupported. We find that biopoesis is possible in $\sim 20\%$ to $\sim 90\%$ of stochastic impact realizations in pessimistic and optimistic scenarios, respectively. This potentially bodes well for life on rocky planets elsewhere because they will have experienced an early episode of enhanced impact bombardment given how planets form.
\end{abstract}

@@

\citet{Benner_2020} argued that the most likely time for the origin of life in an RNA-world scenario would have been $4.36 \pm 0.1$ Ga in the wake of a $\sim 2 \times 10^{22}$ kg ($\sim 2100$ km) asteroid impact called Moneta. They suggested that iron delivered by Moneta's core would have reacted with impact-vaporized ocean water to generate a reducing Hadean atmosphere conducive to the photochemical generation of essential prebiotic nitriles like HCN and HCCCN. Nitriles are required in prebiotic schemes that synthesize ribonucleotide precursors to RNA \citep{Patel_2015,Powner_2009,Becker_2019}. A transient period of profuse nitrile production after a Moneta impact could have been a window of opportunity for the origin of RNA and, ultimately, life.

A single Moneta impact could also explain the highly-siderophile elements (i.e. iron-loving, abbreviated HSEs) in the Moon's and the Earth's mantles. During the Moon-forming impact, HSEs should have been sequestered in each planetary core, so over-abundant HSEs in today's mantles are commonly interpreted as evidence for late-accretion impactors. Earth's mantle has substantially more HSEs than the Moon by an amount that cannot be accounted for by Earth's greater gravitational cross section \citep{Day_2015}. A single massive Moneta impact could explain the Earth-Moon HSE discrepancy because, by the statistics of small numbers, Moneta could have missed the Moon and hit the Earth \citep{Bottke_2010}.

However, the lunar HSE depletion may be explained without a Moneta impact. Lunar HSEs could have been lost to space during impact-delivery because of the Moon's small gravity \citep{Kraus_2015}. Alternatively, HSEs delivered to the Moon during its $\sim 150$ million-year magma ocean could have been sequestered in the core due to iron sulfide exsolution \citep{Morbidelli_2018,Rubie_2016}. Finally, there is some chance that the Earth's HSEs do not record late impacts because the Moon-forming impact delivered HSEs \citep{Sleep_2016}, or because HSEs are gradual core contributions over time from mantle plumes \citep{Halliday_2023,Mundl_2020}. If indeed Earth's HSEs reflect asteroid bombardment after the Moon formed, then the HSEs can be explained by multiple $\sim 500$ to 2000 km impacts rather than a single big ($\sim 2100$ km) collision.

Recently, \citet{Wogan_2023} used photochemical models of post-impact atmospheres to show that impacts significantly smaller than Moneta can efficiently produce prebiotic molecules. In their simulations, impactor iron equilibrates with vaporized ocean water to generate atmospheric H$_2$, CH$_4$ and NH$_3$ that form thermochemically as the reducing steam atmosphere cools. Once steam condenses to an ocean, subsequent photochemistry of a Titan-like, hazy atmosphere produces prebiotic nitriles. Modeling shows that the production of both HCN and HCCCN only occurs when $\mathrm{CH_4} / \mathrm{CO_2} \gtrsim 0.1$ while the atmosphere is hazy. In an optimistic scenario, which includes nickel-catalyzed methane production, modeling suggests that $\mathrm{CH_4} / \mathrm{CO_2} > 0.1$ (i.e. significant nitriles) occurs for impacts $> 4 \times 10^{20}$ kg ($> 570$ km diameter). In a pessimistic modeling case, which assumes only a fraction of impactor iron reacts with the atmosphere \citep{Citron_2022}, a $> 5 \times 10^{21}$ kg ($> 1330$ km diameter) impactor is required to produce substantial HCN and HCCCN. Regardless of uncertainty, \citet{Wogan_2023} show that an impactor much smaller than Moneta could deliver the nitriles needed for an RNA origin of life.

Here, we use the \citet{Wogan_2023} results, along with Monte-Carlo simulations of Earth's impact history to make an alternative estimate for when life most likely emerged during an RNA-first process. Our calculations account for the possibility of planet sterilization by impacts that vaporize the ocean \citep{Sleep_1989}. We assume that a life-starting impact is one that produces significant prebiotic nitriles and is not subsequently followed by an ocean-vaporizing impact that destroys the biosphere. By considering the fraction of stochastic impact realizations that do not have an life-starting impact, we also estimate the probability of life beginning if Earth history was rerun.

@@

We assume that the size-frequency distribution of impactors is identical to the main-belt asteroids \citep[Extended Data Figure 1 in][]{Marchi_2014}. Data for the frequency of main-belt asteroids only extends down to about 1 km diameter ($1.3 \times 10^{12} $ kg). To extend the distribution to smaller objects, we use the observed 1400 ratio between the frequency of $> 1$ km and $> 20$ km craters on the Moon following \citet{Morbidelli_2018}. Crater scaling relations suggest that 1 km and 20 km craters on the Moon corresponds to 50 m and 1 km asteroids, respectively \citep{Morbidelli_2018}. The largest main belt asteroid is $\sim 1000$ km, so we log-linearly extrapolate to larger impactors following \citet{Marchi_2014} (Figure \ref{fig:sfd_and_flux}). Finally, we normalize the size-frequency distribution to the impact mass required to make a 1 km crater on the moon (50 m object, $m_0 = 1.64 \times 10^{8}$ kg). Figure \ref{fig:sfd_and_flux}a shows the resulting size-frequency distribution.

\begin{figure}
  \centering
  \includegraphics[width=1.0\textwidth]{figures/SFD_and_flux.pdf}
  \caption{Our assumed (a) size-frequency distribution of impactors and (b) lunar cratering record. Most of the size-frequency distribution is identical to the main-belt asteroids \citep[Extended Data Figure 1 in][]{Marchi_2014}. We extrapolate to objects $\lesssim 10^{11}$ kg using the observed frequency ratio of $> 1$ km and $> 20$ km lunar craters following \citet{Morbidelli_2018}, and log-linearly extrapolate to asteroids $\gtrsim 10^{21}$ kg following \citet{Marchi_2014}. The lunar cratering record is the red line in Figure 5 of \citet{Morbidelli_2018}.}
  \label{fig:sfd_and_flux}
\end{figure}

The flux of impactors, $F_0(t)$, can be estimated from the lunar cratering record. We adopt the accretionary tail scenario discussed in \citet{Morbidelli_2018}. In other words, we assume that the ``late heavy bombardment'' did not happen \citep{Cartwright_2022,Hartmann_2019,Zellner_2017}, and that the impact flux on Earth was constantly decreasing throughout the Hadean eon. Figure \ref{fig:sfd_and_flux}b shows $n_m$, our assumed lunar impact history taken from \citet{Morbidelli_2018} (red line in their Figure 5). We use the parameterization $n_m = 10^{a e^{b t} + c}$, where $a$, $b$, and $c$ are fit parameters given in Figure \ref{fig:sfd_and_flux}b.

Extrapolating the lunar impact history to Earth requires correcting for Earth's greater gravitational attraction. Assuming that the approach velocity of impactors far from the Earth and Moon was on average 18 km/s \citep{Morbidelli_2018}, then Earth should receive $s_0 = 1.36$ times more impacts per surface area than the Moon. Therefore, the number of impacts on Earth is

\begin{equation}
  N_0 = A_\oplus s_0 n_m = A_\oplus s_0 10^{a e^{b t} + c}
  \label{eq:num_earth_impacts}
\end{equation}
Equation \eqref{eq:num_earth_impacts} also accounts for Earth's surface area ($A_\oplus$), making $N_0$ the number of impacts on Earth since age $t$ that cause a crater $> 1$ km on the moon. As discussed previously, \citet{Morbidelli_2018} use crater scaling relations to show that a 1 km lunar crater corresponds to a 50 m object with mass $m_0 = 1.64 \times 10^8$ kg.

The time derivative of $N_0$ gives the flux, $F_0$:

@@

To simulate impact histories, we consider a grid of $s\sim 200$ impactor masses between $10^{12}$ and $10^{23}$ kg, and a grid of $\sim 60$ times between 4.5 and 3.5 Ga. Indexes $j$ and $i$ indicate the mass and time grid cells respectively, while, for example, $j-\frac{1}{2}$ and $j+\frac{1}{2}$ indicate the edges of the grid cell. We can compute the expected number of impacts within a mass and time grid cell with the following

\begin{equation}
  \overline{N}_{ij} = N(m_{j-\frac{1}{2}},t_{i-\frac{1}{2}},t_{i+\frac{1}{2}}) - N(m_{j+\frac{1}{2}},t_{i-\frac{1}{2}},t_{i+\frac{1}{2}})
\end{equation}
Next, we sample a Poisson distribution for each $\overline{N}_{ij}$, giving a stochastic number of impacts in each mass and time grid cell which constitutes an impact history. By modeling impacts as a Poisson process, we are assuming that each impact does not change the probability of the next one (i.e., impacts are independent). Performing this sampling thousands of times captures many of the possible impact histories. We only keep impact histories that accrete a total mass of $2 \times 10^{22}$ and $6 \times 10^{22}$ kg because this is approximately the mass implied by the HSEs in Earth's mantle, assuming that HSEs were contributed by late accretion. This range of masses is based on \citet{Day_2015} who used mantle HSEs to suggest that the Earth was impacted by $\sim 3 \times 10^{22}$ to $4.8 \times 10^{22}$ kg of material, but we choose a wider range of masses because HSEs could have been lost to Earth's core or to space during impacts \citep{Marchi_2018}.

Our final step is to assign impact velocities to each collision in the many sampled impact histories. Appendix Figure \ref{fig:velocity_distribution} shows Earth's impact velocity distribution. We created this distribution using the JPL database of close approaches to Earth by small bodies \citep{Park_2023}, considering all close approach asteroids within 0.05 AU to Earth. The database gives the approach velocity of each asteroid far from Earth, so we compute the impact velocity by accounting for Earth's gravitational potential energy.

The final collection of sampled impact histories can then be used to compute impact statistics that are relevant to the timing and likelihood of the origin of life.

@@

Moneta also occurs in the Figure \ref{fig:example_impact_histories}b impact realization. In this scenario, any life created in the wake of Moneta will not be subsequently impact-exterminated because Moneta is the last impact to vaporize the ocean.

Figure \ref{fig:example_impact_histories}c shows a impact history where Moneta does not occur. Instead, Earth's mantle HSEs are delivered by multiple $10^{21}$ to $4 \times 10^{21}$ kg collisions. Whether this bombardment history would cause biopoesis depends on the required minimum impact mass to produce significant origin of life nitriles (e.g. HCN and HCCCN). \citet{Wogan_2023} uses photochemical models of post-impact atmospheres to show that, optimistically, an impact $> 4 \times 10^{20}$ kg should give rise to an atmosphere that makes significant HCN and HCCCN. Under this optimistic requirement, the last ocean-vaporizing impact in Figure \ref{fig:example_impact_histories}c could successfully start life and occurs at 4.38 Ga with mass $3 \times 10^{21}$ kg. However, with pessimistic modeling assumptions, \citet{Wogan_2023} finds that a $> 5 \times 10^{21}$ kg impactor is instead required for substantial nitrile delivery to the surface. In this alternative scenario, an origin of life never occurs in Figure \ref{fig:example_impact_histories}c because there is no impactor larger than $5 \times 10^{21}$ kg.

@@

As illustrated in Figure \ref{fig:example_impact_histories}, a life-starting impact does not occur in every simulated impact history. In our impact-driven model for the origin of life, biopoesis requires (1) an impact of sufficient mass to make prebiotic nitriles and (2) the lack of a later ocean-vaporizing collision that destroys the biosphere without rekindling it. Figure \ref{fig:probabilities_of_impacts} shows the probability of both conditions occurring as a function of the minimum impact mass that produces prebiotic molecules. For example, consider the \citet{Wogan_2023} optimistic minimum mass of $4 \times 10^{20}$ kg indicated in Figure \ref{fig:probabilities_of_impacts}. All simulated impact histories have at least one impact bigger than $4 \times 10^{20}$ kg (blue line in Figure \ref{fig:probabilities_of_impacts} is 1.0 at the given mass). However, the probability of the origin of life in this scenario is 92\%, because 8\% of the time the last $4 \times 10^{20}$ kg collision is followed by a smaller ocean-vaporizing impact that perhaps sterilizes the planet (orange dashed line in Figure \ref{fig:probabilities_of_impacts} with 0.92 at $4 \times 10^{20}$ kg). An origin of life is less probable if we consider the \citet{Wogan_2023} pessimistic minimum mass ($> 5 \times 10^{21}$ kg) to produce significant origin of life nitriles. A $> 5 \times 10^{21}$ kg impact without later ocean vaporization occurs 41\% of the time. A life-starting impact was therefore relatively likely on the early Earth, and was strongly favored (92\% probability) under optimistic assumptions.

\begin{figure}
  \centering
  \includegraphics[width=0.85\textwidth]{figures/timing_and_mass.pdf}
  \caption{The timing and mass of a life-starting impact on the early Earth. (a) considers an optimistic minimum impact mass needed to generate the molecules for an RNA origin of life and (b) examines a pessimistic minimum impact mass \citep{Wogan_2023}. In either (a) or (b), panels (i) and (ii) show probability distributions for the timing and mass, respectively, of the last impact the produces prebiotic molecules which does not have a later ocean-vaporizing collision. The vertical dashed lines on (a) i and (b) i plot are median values, and the vertical dotted lines are the 95\% confidence intervals. In an impact-driven origin of life, biopoesis most likely occurred at $4.34$ Ga (95\% CI: 3.89 to 4.46 Ga) assuming panel (a) and at $4.36$ Ga (95\% CI: 3.94 to 4.49 Ga) assuming panel (b).}
  \label{fig:time_and_mass}
\end{figure}

To estimate the most likely timing for the origin of life, we consider both the \citet{Wogan_2023} optimistic and pessimistic criteria. Figure \ref{fig:time_and_mass}a is the optimistic case, showing the timing (Figure \ref{fig:time_and_mass}a (i)) and mass (Figure \ref{fig:time_and_mass}a (ii)) of the last $> 4 \times 10^{20}$ kg impactor for the 92\% of impact histories that do not experience later ocean vaporization. The median timing of a life-starting impact is 4.34 Ga with 95\% uncertainty between 3.89 and 4.46 Ga. The large uncertainty mostly stems from the stochastic, Poisson nature of asteroid bombardment. The probability distribution for the mass of the life-starting impact peaks at $4 \times 10^{20}$ kg and decreases with increasing mass because of the size-frequency distribution of asteroids (Figure \ref{fig:sfd_and_flux}a). There is only a 10\% chance that the impactor is Moneta-sized (i.e., between $2 \times 10^{22}$ and $6 \times 10^{22}$ kg).

@@

In this model for biopoesis, life would probably start approximately within 10s of millions of years after a life-starting impact because this is the maximum duration of the H$_2$- and CH$_4$- rich atmosphere that makes nitriles like HCN \citep{Wogan_2023}. We ignore the $< 10$s million year span between an impact and life's origin because it is small compared to our $\sim 500$ million-year uncertainty for the timing of a life-starting impact.

@@

Our estimated timing for biopoesis is compatible with the earliest well-accepted geologic evidence of life in the form of stromatolites in the 3.5 Ga Pilbara block of western Australia \citep{Walter_1980,Buick_1981,Van_2018}. Older geologic evidence of life exists, such as a $> 3.7$ Ga black shale metamorphosed to graphite with negative $\delta^{13}$C characteristic of biology . Also, \citet{Bell_2015} discovered graphite inclusions in a 4.1 Ga zircon grain with $^{13}$C depletions compatible with biological activity, but this is a tentative biosignature because there are abiotic mechanisms for making isotopically light carbon \citep{Javaux_2019}. If the \citet{Bell_2015} finding is indeed a sign of life, then it would be mostly compatible with our estimating timing in Figure \ref{fig:time_and_mass}, because we find that there is only a $\sim 10\%$ chance of life starting after 4.1 Ga.

It is temping to extrapolate our results beyond the early Earth to exoplanets, but we must do so with caution because planet's orbiting different stars may have bombardment histories unlike the Hadean. Planets in the habitable zone of an M-type star during the stellar main-sequence phase were interior to the habitable zone for several hundred million years during the super luminous pre-main-sequence phase \citep{Luger_2015}. \citet{Lichtenberg_2022} uses N-body simulations to show that, for planets hosted by M-dwarfs, large asteroid impacts are not useful for prebiotic chemistry because big impacts most likely occur within 100 million years of planet formation during the stellar pre-main-sequence when the planet is outside of the habitable zone. The pre-main-sequence phase is not an issue for an impact-induced origin of life on planets orbiting sun-like stars because the phase only lasts 10s Myrs, while large asteroid impacts occur for 100s Myrs \citep{Lichtenberg_2022}. Therefore, our result that a life-starting impact was likely on the early Earth (92\% chance in an optimistic case) might be best extrapolated to habitable zone exoplanets orbiting sun-like stars.
